Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R

### 2017-04-21 meehan_et_al_inla_nmix_jss/article.tex
fixed code format to italic

\begin{Code}
R> data4sim <- function(n.sites = 72,   # number study sites
+                       n.surveys = 3,  # number replicate surveys
+                       n.years = 9,    # number years
+                       lam.b0 = 2.0,   # intercept for log lambda
+                       lam.b1 = 2.0,   # slope for log lambda cov 1
+                       lam.b2 = -3.0,  # slope for log lambda cov 2
+                       lam.b4 = 1.0,   # slope for log lambda year
+                       p.b0 = 1.0,     # intercept for logit p
+                       p.b1 = -2.0,    # slope for logit p cov 1
+                       p.b3 = 1.0,     # slope for logit p cov 3
+                       disp.size = 3.0 # size of overdispersion
+                       ){
+  # setup
+  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
+  N.tr <- array(dim = c(n.sites, n.years))           # for abund
+  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex1 counts
+  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex2 counts
+  # abundance covariate values
+  x.lam.1 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  x.lam.2 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  yrs <- 1:n.years
+  yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
+  yr <- array(rep(yrs, each = n.sites), dim = c(n.sites, n.years))
+  # fill abundance array
+  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.b4*yr)
+  for(i in 1:n.sites){
+    for(k in 1:n.years){
+    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)
+  }}
+  # detection covariate values
+  x.p.1 <- array(x.lam.1[,1], dim = c(n.sites, n.surveys, n.years))
+  x.p.3 <- array(as.numeric(scale(runif(n = n.sites*n.surveys*n.years, 
+                                        -0.5, 0.5), scale=F)), 
+                 dim=c(n.sites, n.surveys, n.years))
+  # average x.p.3 per site-year
+  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
+  out1 <- c()
+  for(k in 1:n.years){
+   chunk1 <- x.p.3.mean[,k]
+   chunk2 <- rep(chunk1, n.surveys)
+   out1 <- c(out1, chunk2)
+  }
+  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
+  # fill count array with site-yr x.p.3
+  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y1[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr1[i,j,k])
+  }}}
+  # fill count array with site-surv-yr x.p.3
+  p.tr2 <- plogis(p.b0 + p.b1 * x.p.1 + p.b3 * x.p.3)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y2[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr2[i,j,k])
+  }}}
+  # return data
+  return(list(n.sites = n.sites, n.surveys = n.surveys, n.years = n.years,
+              x.p.1 = x.p.1[,1,1], x.p.3 = x.p.3, x.p.3.mean = x.p.3.mean,
+              x.p.3.arr = x.p.3.arr, x.lam.1 = x.lam.1[,1],
+              x.lam.2 = x.lam.2[,1], yr = yr[1,], y1 = y1, y2 = y2,
+              lam.tr = lam.tr, N.tr = N.tr))
+  } #end function
R> set.seed(12345) # make results reproducible
R> sim.data <- data4sim() # create datasets
\end{Code}

@@

\begin{Code}
R> jags.model.string <- "
+    model {
+    # priors
+    intP ~ dnorm(0, 0.01)       # detection intercept
+    bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
+    bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
+    intLam ~ dnorm(0, 0.01)     # lambda intercept
+    bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
+    bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
+    bYr ~ dnorm(0, 0.01)        # year effect
+    overDisEst ~ dunif(0, 5)    # overdispersion size
+    # abundance component
+    for (k in 1:nYears){
+      for (i in 1:nSites){
+        N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial
+        prob[i, k] <- overDisEst / (overDisEst + lambda[i, k])
+        log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + 
+                             (bCov2Lam * x.lam.2[i]) + (bYr * yr[k])
+    # detection component
+        for (j in 1:nSurveys){
+          y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
+          p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
+          lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, k])
+        } # close j loop
+      } # close i loop
+    } # close k loop
+  } # close model loop
+  "
\end{Code}

@@

\begin{Code}
R> params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
+              "bYr", "overDisEst") # parameters to monitor
R> jags.data <- list(y = sim.data$y1, x.lam.1 = sim.data$x.lam.1,
+               x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr,
+               x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3.mean,
+               nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
+               nYears = sim.data$n.years)
R> N.init <- sim.data$y1 # initial count values
R> N.init[is.na(N.init)] <- 1 # clean up NA's
R> N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
R> inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01),
+               intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01),
+               bCov2Lam = rnorm(1, 0, 0.01), bCov1Lam = rnorm(1, 0, 0.01),
+               bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01),
+               overDisEst = runif(1, 0.5, 2.5))
\end{Code}

@@

\begin{Code}
R> nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10 # run parameters
R> out.jags <- run.jags(model = jags.model.string, data = jags.data,
+                       monitor = params, n.chains = nc, inits = inits,
+                       burnin = nb, adapt = na, sample = ni, thin = nt,
+                       modules = "glm on", method = "parallel")
\end{Code}

@@

\begin{Code}
R> y.unmk <- sim.data$y1[ , , 1]
R> for(i in 2:sim.data$n.years){
+    y.chunk <- sim.data$y1[ , , i]
+    y.unmk <- rbind(y.unmk, y.chunk)
+  } 
R> x.lam.1.unmk <- rep(sim.data$x.lam.1, sim.data$n.years)
R> x.lam.2.unmk <- rep(sim.data$x.lam.2, sim.data$n.years)
R> yr.unmk <- rep(sim.data$yr, each = sim.data$n.sites)
R> x.p.1.unmk <- rep(sim.data$x.p.1, sim.data$n.years)
R> x.p.3.unmk <- c(sim.data$x.p.3.mean)
R> site.covs.unmk <- data.frame(x.lam.1.unmk, x.lam.2.unmk, yr.unmk,
+                               x.p.1.unmk, x.p.3.unmk)
R> unmk.data <- unmarkedFramePCount(y = y.unmk, siteCovs = site.covs.unmk)
\end{Code}

@@

\begin{Code}
R> out.unmk <- pcount(~ x.p.1.unmk + x.p.3.unmk
+                     ~ x.lam.1.unmk + x.lam.2.unmk + yr.unmk,
+                       data = unmk.data, mixture = "NB")
\end{Code}

@@

\begin{Code}

R> y.inla <- sim.data$y1[ , , 1]
R> for(i in 2:sim.data$n.years){
+    y.chunk <- sim.data$y1[ , , i]
+    y.inla <- rbind(y.inla, y.chunk)
+  }
R> x.lam.1.inla <- rep(sim.data$x.lam.1, sim.data$n.years)
R> x.lam.2.inla <- rep(sim.data$x.lam.2, sim.data$n.years)
R> yr.inla <- rep(sim.data$yr, each = sim.data$n.sites)
R> x.p.1.inla <- rep(sim.data$x.p.1, sim.data$n.years)
R> x.p.3.inla <- c(sim.data$x.p.3.mean)
R> counts.and.count.covs <- inla.mdata(y.inla, 1, x.lam.1.inla, 
+                                      x.lam.2.inla, yr.inla)
\end{Code}

@@

\begin{Code}
R> out.inla <- inla(counts.and.count.covs ~ 1 + x.p.1.inla + x.p.3.inla,
+           data = list(counts.and.count.covs = counts.and.count.covs,
+                       x.p.1.inla = x.p.1.inla, x.p.3.inla = x.p.3.inla),
+           family = "nmixnb",
+           control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
+                                prec.intercept = 0.01),
+           control.family = list(hyper = 
+                                  list(theta1 = list(param = c(0, 0.01)),
+                                       theta2 = list(param = c(0, 0.01)),
+                                       theta3 = list(param = c(0, 0.01)),
+                                       theta4 = list(param = c(0, 0.01)),
+                                       theta5 = list(prior = "flat",
+                                                     param = numeric()))))
\end{Code}

@@

\begin{Code}
R> data("mallard") # get mallard data from unmarked package
R> mallard.y.unmk <- mallard.y # format unmarked data
R> mallard.site.unmk <- mallard.site
R> mallard.obs.unmk <- mallard.obs
R> mallardUMF <- unmarkedFramePCount(mallard.y.unmk, 
+                                    siteCovs = mallard.site.unmk,
+                                    obsCovs = mallard.obs.unmk)
R> mallard.out.unmk <- pcount(~ ivel+ date + I(date^2)
+                             ~ length + elev + forest,
+                               mixture = "NB", mallardUMF, K=30) 
\end{Code}

@@

\begin{Code}
R> mallard.y.inla <- mallard.y # format inla data
R> mallard.length.inla <- mallard.site[,2]
R> mallard.elev.inla <- mallard.site[,1]
R> mallard.forest.inla <- mallard.site[,3]
R> mallard.ivel.inla <- rowMeans(mallard.obs$ivel, na.rm=T) # site average
R> mallard.ivel.inla[is.na(mallard.ivel.inla)] <- 
+                                       mean(mallard.ivel.inla, na.rm = T)
R> mallard.date.inla <- rowMeans(mallard.obs$date, na.rm=T) # site average
R> mallard.date2.inla <- mallard.date.inla^2
R> counts.and.count.covs <- inla.mdata(mallard.y.inla, 1, 
+                                      mallard.length.inla,
+                                      mallard.elev.inla, 
+                                      mallard.forest.inla)
R> mallard.out.inla <- inla(counts.and.count.covs ~ 1 + mallard.ivel.inla +
+                           mallard.date.inla + mallard.date2.inla,
+      data = list(counts.and.count.covs = counts.and.count.covs,
+                           mallard.ivel.inla = mallard.ivel.inla,
+                           mallard.date.inla = mallard.date.inla,
+                           mallard.date2.inla = mallard.date2.inla),
+      family = "nmixnb",
+      control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
+                               prec.intercept = 0.01),
+      control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
+                                    theta2 = list(param = c(0, 0.01)),
+                                    theta3 = list(param = c(0, 0.01)),
+                                    theta4 = list(param = c(0, 0.01)),
+                                    theta5 = list(prior = "flat",
+                                                  param = numeric()))))
\end{Code}

@@

The key observation is that both the Poisson and the binomial distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{binomial}(y; n, p) = \text{binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-binomial product
\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
\begin{eqnarray}
    \text{Prob}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-liklihood can then be evaluated using the following simple \proglang{R}-code.
\begin{Code}
R> fac <- 1; ff <- lambda * (1-p)
R> for(i in (n.max - y):1){
+    fac <- 1 + fac * ff / i
+  } 
R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
+           log(fac)
\end{Code}
Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small. Note that we are computing \texttt{fac} starting with the smallest contributions, which are more numerically stable.

>>>>

\begin{CodeInput}
R> data4sim <- function(n.sites = 72,   # number study sites
+                       n.surveys = 3,  # number replicate surveys
+                       n.years = 9,    # number years
+                       lam.b0 = 2.0,   # intercept for log lambda
+                       lam.b1 = 2.0,   # slope for log lambda cov 1
+                       lam.b2 = -3.0,  # slope for log lambda cov 2
+                       lam.b4 = 1.0,   # slope for log lambda year
+                       p.b0 = 1.0,     # intercept for logit p
+                       p.b1 = -2.0,    # slope for logit p cov 1
+                       p.b3 = 1.0,     # slope for logit p cov 3
+                       disp.size = 3.0 # size of overdispersion
+                       ){
+  # setup
+  if(n.years %% 2 == 0) {n.years <- n.years + 1}     # make years odd
+  N.tr <- array(dim = c(n.sites, n.years))           # for abund
+  y1 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex1 counts
+  y2 <- array(dim = c(n.sites, n.surveys, n.years))  # for ex2 counts
+  # abundance covariate values
+  x.lam.1 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  x.lam.2 <- array(as.numeric(scale(runif(n = n.sites, -0.5, 0.5), 
+                   scale = F)), dim = c(n.sites, n.years))
+  yrs <- 1:n.years
+  yrs <- (yrs - mean(yrs)) / (max(yrs - mean(yrs))) / 2
+  yr <- array(rep(yrs, each = n.sites), dim = c(n.sites, n.years))
+  # fill abundance array
+  lam.tr <- exp(lam.b0 + lam.b1*x.lam.1 + lam.b2*x.lam.2 + lam.b4*yr)
+  for(i in 1:n.sites){
+    for(k in 1:n.years){
+    N.tr[i, k] <- rnbinom(n = 1, mu = lam.tr[i, k], size = disp.size)
+  }}
+  # detection covariate values
+  x.p.1 <- array(x.lam.1[,1], dim = c(n.sites, n.surveys, n.years))
+  x.p.3 <- array(as.numeric(scale(runif(n = n.sites*n.surveys*n.years, 
+                                        -0.5, 0.5), scale=F)), 
+                 dim=c(n.sites, n.surveys, n.years))
+  # average x.p.3 per site-year
+  x.p.3.mean <- apply(x.p.3, c(1,3), mean, na.rm=F)
+  out1 <- c()
+  for(k in 1:n.years){
+   chunk1 <- x.p.3.mean[,k]
+   chunk2 <- rep(chunk1, n.surveys)
+   out1 <- c(out1, chunk2)
+  }
+  x.p.3.arr <- array(out1, dim=c(n.sites, n.surveys, n.years))
+  # fill count array with site-yr x.p.3
+  p.tr1 <- plogis(p.b0 + p.b1*x.p.1 + p.b3*x.p.3.arr)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y1[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr1[i,j,k])
+  }}}
+  # fill count array with site-surv-yr x.p.3
+  p.tr2 <- plogis(p.b0 + p.b1 * x.p.1 + p.b3 * x.p.3)
+  for (i in 1:n.sites){
+    for (k in 1:n.years){
+      for (j in 1:n.surveys){
+        y2[i,j,k] <- rbinom(1, size = N.tr[i,k], prob = p.tr2[i,j,k])
+  }}}
+  # return data
+  return(list(n.sites = n.sites, n.surveys = n.surveys, n.years = n.years,
+              x.p.1 = x.p.1[,1,1], x.p.3 = x.p.3, x.p.3.mean = x.p.3.mean,
+              x.p.3.arr = x.p.3.arr, x.lam.1 = x.lam.1[,1],
+              x.lam.2 = x.lam.2[,1], yr = yr[1,], y1 = y1, y2 = y2,
+              lam.tr = lam.tr, N.tr = N.tr))
+  } #end function
R> set.seed(12345) # make results reproducible
R> sim.data <- data4sim() # create datasets
\end{CodeInput}

@@

\begin{CodeInput}
R> jags.model.string <- "
+    model {
+    # priors
+    intP ~ dnorm(0, 0.01)       # detection intercept
+    bCov1P ~ dnorm(0, 0.01)     # detection cov 1 effect
+    bCov3P ~ dnorm(0, 0.01)     # detection cov 3 effect
+    intLam ~ dnorm(0, 0.01)     # lambda intercept
+    bCov1Lam ~ dnorm(0, 0.01)   # lambda cov 1 effect
+    bCov2Lam ~ dnorm(0, 0.01)   # lambda cov 2 effect
+    bYr ~ dnorm(0, 0.01)        # year effect
+    overDisEst ~ dunif(0, 5)    # overdispersion size
+    # abundance component
+    for (k in 1:nYears){
+      for (i in 1:nSites){
+        N[i, k] ~ dnegbin(prob[i, k], overDisEst) # negative binomial
+        prob[i, k] <- overDisEst / (overDisEst + lambda[i, k])
+        log(lambda[i, k]) <- intLam + (bCov1Lam * x.lam.1[i]) + 
+                             (bCov2Lam * x.lam.2[i]) + (bYr * yr[k])
+    # detection component
+        for (j in 1:nSurveys){
+          y[i, j, k] ~ dbin(p[i,j,k], N[i,k])
+          p[i, j, k] <- exp(lp[i,j,k]) / (1 + exp(lp[i,j,k]))
+          lp[i, j, k] <- intP + (bCov1P * x.p.1[i]) + (bCov3P * x.p.3[i, k])
+        } # close j loop
+      } # close i loop
+    } # close k loop
+  } # close model loop
+  "
\end{CodeInput}

@@

\begin{CodeInput}
R> params <- c("intP", "bCov1P", "bCov3P", "intLam", "bCov1Lam","bCov2Lam",
+              "bYr", "overDisEst") # parameters to monitor
R> jags.data <- list(y = sim.data$y1, x.lam.1 = sim.data$x.lam.1,
+               x.lam.2 = sim.data$x.lam.2, yr = sim.data$yr,
+               x.p.1 = sim.data$x.p.1, x.p.3 = sim.data$x.p.3.mean,
+               nSites = sim.data$n.sites, nSurveys = sim.data$n.surveys,
+               nYears = sim.data$n.years)
R> N.init <- sim.data$y1 # initial count values
R> N.init[is.na(N.init)] <- 1 # clean up NA's
R> N.init <- apply(N.init, c(1, 3), max) + 1 # zero values cause trouble
R> inits <- function() list(N = N.init, intLam = rnorm(1, 0, 0.01),
+               intP = rnorm(1, 0, 0.01), bCov1P = rnorm(1, 0, 0.01),
+               bCov2Lam = rnorm(1, 0, 0.01), bCov1Lam = rnorm(1, 0, 0.01),
+               bCov3P = rnorm(1, 0, 0.01), bYr = rnorm(1, 0, 0.01),
+               overDisEst = runif(1, 0.5, 2.5))
\end{CodeInput}

@@

\begin{CodeInput}
R> nc <- 3; na <- 2500; nb <- 2500; ni <- 5000; nt <- 10 # run parameters
R> out.jags <- run.jags(model = jags.model.string, data = jags.data,
+                       monitor = params, n.chains = nc, inits = inits,
+                       burnin = nb, adapt = na, sample = ni, thin = nt,
+                       modules = "glm on", method = "parallel")
\end{CodeInput}

@@

\begin{CodeInput}
R> y.unmk <- sim.data$y1[ , , 1]
R> for(i in 2:sim.data$n.years){
+    y.chunk <- sim.data$y1[ , , i]
+    y.unmk <- rbind(y.unmk, y.chunk)
+  } 
R> x.lam.1.unmk <- rep(sim.data$x.lam.1, sim.data$n.years)
R> x.lam.2.unmk <- rep(sim.data$x.lam.2, sim.data$n.years)
R> yr.unmk <- rep(sim.data$yr, each = sim.data$n.sites)
R> x.p.1.unmk <- rep(sim.data$x.p.1, sim.data$n.years)
R> x.p.3.unmk <- c(sim.data$x.p.3.mean)
R> site.covs.unmk <- data.frame(x.lam.1.unmk, x.lam.2.unmk, yr.unmk,
+                               x.p.1.unmk, x.p.3.unmk)
R> unmk.data <- unmarkedFramePCount(y = y.unmk, siteCovs = site.covs.unmk)
\end{CodeInput}

@@

\begin{CodeInput}
R> out.unmk <- pcount(~ x.p.1.unmk + x.p.3.unmk
+                     ~ x.lam.1.unmk + x.lam.2.unmk + yr.unmk,
+                       data = unmk.data, mixture = "NB")
\end{CodeInput}

@@

\begin{CodeInput}

R> y.inla <- sim.data$y1[ , , 1]
R> for(i in 2:sim.data$n.years){
+    y.chunk <- sim.data$y1[ , , i]
+    y.inla <- rbind(y.inla, y.chunk)
+  }
R> x.lam.1.inla <- rep(sim.data$x.lam.1, sim.data$n.years)
R> x.lam.2.inla <- rep(sim.data$x.lam.2, sim.data$n.years)
R> yr.inla <- rep(sim.data$yr, each = sim.data$n.sites)
R> x.p.1.inla <- rep(sim.data$x.p.1, sim.data$n.years)
R> x.p.3.inla <- c(sim.data$x.p.3.mean)
R> counts.and.count.covs <- inla.mdata(y.inla, 1, x.lam.1.inla, 
+                                      x.lam.2.inla, yr.inla)
\end{CodeInput}

@@

\begin{CodeInput}
R> out.inla <- inla(counts.and.count.covs ~ 1 + x.p.1.inla + x.p.3.inla,
+           data = list(counts.and.count.covs = counts.and.count.covs,
+                       x.p.1.inla = x.p.1.inla, x.p.3.inla = x.p.3.inla),
+           family = "nmixnb",
+           control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
+                                prec.intercept = 0.01),
+           control.family = list(hyper = 
+                                  list(theta1 = list(param = c(0, 0.01)),
+                                       theta2 = list(param = c(0, 0.01)),
+                                       theta3 = list(param = c(0, 0.01)),
+                                       theta4 = list(param = c(0, 0.01)),
+                                       theta5 = list(prior = "flat",
+                                                     param = numeric()))))
\end{CodeInput}

@@

\begin{CodeInput}
R> data("mallard") # get mallard data from unmarked package
R> mallard.y.unmk <- mallard.y # format unmarked data
R> mallard.site.unmk <- mallard.site
R> mallard.obs.unmk <- mallard.obs
R> mallardUMF <- unmarkedFramePCount(mallard.y.unmk, 
+                                    siteCovs = mallard.site.unmk,
+                                    obsCovs = mallard.obs.unmk)
R> mallard.out.unmk <- pcount(~ ivel+ date + I(date^2)
+                             ~ length + elev + forest,
+                               mixture = "NB", mallardUMF, K=30) 
\end{CodeInput}

@@

\begin{CodeInput}
R> mallard.y.inla <- mallard.y # format inla data
R> mallard.length.inla <- mallard.site[,2]
R> mallard.elev.inla <- mallard.site[,1]
R> mallard.forest.inla <- mallard.site[,3]
R> mallard.ivel.inla <- rowMeans(mallard.obs$ivel, na.rm=T) # site average
R> mallard.ivel.inla[is.na(mallard.ivel.inla)] <- 
+                                       mean(mallard.ivel.inla, na.rm = T)
R> mallard.date.inla <- rowMeans(mallard.obs$date, na.rm=T) # site average
R> mallard.date2.inla <- mallard.date.inla^2
R> counts.and.count.covs <- inla.mdata(mallard.y.inla, 1, 
+                                      mallard.length.inla,
+                                      mallard.elev.inla, 
+                                      mallard.forest.inla)
R> mallard.out.inla <- inla(counts.and.count.covs ~ 1 + mallard.ivel.inla +
+                           mallard.date.inla + mallard.date2.inla,
+      data = list(counts.and.count.covs = counts.and.count.covs,
+                           mallard.ivel.inla = mallard.ivel.inla,
+                           mallard.date.inla = mallard.date.inla,
+                           mallard.date2.inla = mallard.date2.inla),
+      family = "nmixnb",
+      control.fixed = list(mean = 0, mean.intercept = 0, prec = 0.01,
+                               prec.intercept = 0.01),
+      control.family = list(hyper = list(theta1 = list(param = c(0, 0.01)),
+                                    theta2 = list(param = c(0, 0.01)),
+                                    theta3 = list(param = c(0, 0.01)),
+                                    theta4 = list(param = c(0, 0.01)),
+                                    theta5 = list(prior = "flat",
+                                                  param = numeric()))))
\end{CodeInput}

@@

The key observation is that both the Poisson and the binomial distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{binomial}(y; n, p) = \text{binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-binomial product
\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
\begin{eqnarray}
    \text{Prob}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-liklihood can then be evaluated using the following simple \proglang{R}-code.
\begin{CodeInput}
R> fac <- 1; ff <- lambda * (1-p)
R> for(i in (n.max - y):1){
+    fac <- 1 + fac * ff / i
+  } 
R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
+           log(fac)
\end{CodeInput}
Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small. Note that we are computing \texttt{fac} starting with the smallest contributions, which are more numerically stable.
